---
title: "BWG"
author: "Nathan Wyatt"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Welcome to week 1 of the Bioinformatics Working Group. My goal for this first
meetup is to cover:
  - some SciNet Basics
  - An introdcution to R/Rstudio 
  - Demonstrate the utility of ChatGPT for learning 

# SciNet: What is it and how to get started
SciNet is the term applied to the high performance computing cluster that the 
USDA has spent a significant amount of money developing. 

It consists of two separate but connected clusters:
  - Ceres - located in Ames Iowa
  - Atlas - located in Starkville Mississippi
  
I primarily use Ceres. There are some subtle differences between the inner 
workings of Ceres and Atlas but in large part they are similar and differences
are well documented.

All SciNet documents can be found at: https://scinet.usda.gov/

## Who can use SciNet?
The short answer is everyone. SciNet does not requrie USDA credentials and so
it is already available for everyone in this room. However, if you do not have 
a USDA email address the access must be requested by your collaborator or SY.

The link to request an account can be found here: https://scinet.usda.gov/about/signup

## How should I log in to SciNet?
The answer is a program called Putty... and you will need local IT to install it. 

## Why should you use SciNet?
In general, SciNet provides a computational space for us all to work without the 
need for IT staff to install and direct our software usage. This means that anything
you want to accomplish is possible as long as you have the know-how. SciNet also
comes with a ton of pre-installed programs that are common to bioinformatics.
  Examples:
    ncbi-blast+ (for blasting)
    samtools (for sequence alignment files)
    bcftools (variant calling and manipulation)
    bedtools (the swiss army knife of genomics)
    seqtk (sequence manipulations)
    GATK (variant calling)
    mummer (whole genome alignment)
    spades (short read genome assembler)

# Moving over to R and Rstudio
R is my prefered language for data science given the large community of 
developers and the ease of reading the code. It is highly flexible and scaleable. 

## Install
You will need the help of IT to install R and Rstudio...

## Getting data into Rstudio so that you can explore
As scientist we tend to store data in excel sheets in ever more complex formats
as we work to make sense of the data. I'd like to present an alternative to this
in the form of Tidy data so that we can maintain a consistent data format yet 
have the flexibility to interogate data. 

If you want to get into the deep end of the pool with R and data science 
check out: https://r4ds.had.co.nz/

But before we get to data lets get setup in R.

## Setup: Libraries
One of the best parts of R is that it comes with a mechanism to install other 
folks perbuilt software and these are usually refered to as 'packages' or 'libriaries'.
As an example, the library 'tidyverse' is a composite package of many data science
softwares. 

To install a library and load it into your working environment:
```{r, include=FALSE}
#if the library is not installed run:
#install.packages(tidyverse)
#load the installed library:
library(tidyverse)
library(janitor)
library(rmarkdown)
library(plotly)
```
## Setup: data
Often loading data into R is the first issue folks need to overcome. It's not an
intuitive process but once practiced becomes second nature. 

First and foremost you need to set a working directory or at least check that you
are in the right place. 

```{r}
###command to figure out where you are
getwd()
###command to set a new location
#setwd()
```
There are numerous functions to read in data but in general the three most common 
will be read.csv() read.table() and read.excel(). Bear in mind that read.excel comes
from the package readxl that would need to be installed and loaded as shown previously.
I typically work with csv and txt files so read.csv() is my go to. 

I've set up some weather data that we will quickly take a look at.

```{r}
#read in the file and label it as "Fargo2022"
Fargo2022 <- read.csv(file = "data/NDAWN_23_2022.csv")
#view the top of the data
paged_table(Fargo2022[1:10,1:10])
```

## Tidy data
Simply put, tidy data refers to structuring data so that each observation is 
in a row, each variable to track is in a column, and values are stored at the 
intersection of rows and columns. 

Is the data we loaded up Tidy?

```{r}
#adding a few options to improve the data and make it Tidy
Fargo2022 <- read.csv(file = "data/NDAWN_23_2022.csv", skip = 1, sep = ",", header = T, row.names = 1)
#quick check data
paged_table(Fargo2022[1:5,1:5])
```
```{r}
Fargo2022 <- filter(Fargo2022, date != "Date")
as.data.frame(sapply(Fargo2022,class))
Fargo2022$Date <- lubridate::ymd(Fargo2022$Date)
as.data.frame(sapply(Fargo2022,class))
Fargo2022 <- clean_names(Fargo2022)
paged_table(Fargo2022[1:5,1:5])
```

## Now that we have Tidy data we can start generating some summary plots

You can also embed plots, for example:

```{r, echo=FALSE}
January2022 <- Fargo2022 %>% filter(date < '2022-02-01')
paged_table(January2022)

Jplot <- January2022 %>% group_by(date) %>%
  summarise(mean_temp = mean(as.numeric(avg_air_temp_f))) %>%
  ggplot(aes(x = date, y = mean_temp)) +
  geom_point() +
  geom_line()

ggplotly(Jplot)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
